Capabilities, MMIO and Robust Safety

### 2020-03-31 capability-notes/driver-notes/main.tex
Some minor notational fixes

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\X}[1]{\ensuremath{\mathrm{#1}}}
\newcommand{\V}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\I}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\T}[1]{\texttt{#1}}
\newcommand{\SL}{Separation Logic\xspace}
\newcommand{\pure}[1]{\tensor[^{\ulcorner}]{#1{}}{^{\urcorner}}} %Hacky {} to avoid double superscript

@@

\newcommand{\MMIO}{\textlog{MMIO}\xspace}

\DeclareMathOperator{\initOKo}{init_{OK}}
\DeclareMathOperator{\initOKos}{init^{\X{sem}}_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\newcommandx{\initOKS}[4][2=r,3=m,4=W]{\initOKos(#1,#2,#3,#4)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\DeclareMathOperator{\initBCos}{init^{\X{sem}}_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
\newcommandx{\initBCS}[3][1=r,2=m,3=W]{\initBCos(#1,#2,#3)}
\DeclareMathOperator{\driverC}{code}
\newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}

\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
\newcommand{\app}{\mathbin{+\!\!+}}

@@

\begin{mathpar}
  \mprset{vskip=0.3em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
    (C[e]);(\emptyset, \textlog{OK}) \longrightarrow^* T';(h',g')
  }
  {g' = \textlog{OK}}
\end{mathpar}

For any closed expression $e$, if $e$ has been verified to only return low
values, then running $e$ wrapped in an adversarial context $C$ from an initial
state, then we can observe that every reachable state is good
($g' = \textlog{OK}$ means that no assertion has failed in $e$).

$C \in AdvCtx$ means that $C$ cannot contain assertions (otherwise one could
trivially contradict the theorem by taking
$C[\cdot] = \textlog{assert false}$), and cannot contain references to raw
memory locations (otherwise one could directly access $e$'s private state,
invalidating the local state encapsulation mechanisms).

\newcommand{\lift}[2]{\textlog{lift}\; #1 \; #2}
\newcommand{\liftP}[1]{\lift{\Psi}{#1}}

$\textlog{lowval}\; x$ intuitively means that $x$ cannot be used to directly
access private (or ``high'') locations. It is formally defined using a logical
relation ``$\liftP{v}$'', which more generally asserts that the value $v$ only
gives direct access to locations that satisfy the predicate $\Psi$.

\[
  \begin{array}{lcl}
    \liftP (\textlog{rec}\; f\; x.\; e) & \eqdef
    & \later \forall v.\; \hoare{\liftP{v}}{e[v/x,\textlog{rec}\;f\;x.\; e/f]}{y.\; \liftP y} \\
    %
    \liftP (v_{1}, v_{2}) & \eqdef & \later (\liftP v_{1}, \liftP v_{2}) \\
    %
    \color{BrickRed} \liftP \ell & \color{BrickRed} \eqdef & \color{BrickRed} \Psi \; \ell \\
    %
    \ldots
  \end{array}
\]

Then, $\textlog{lowval}\; x$ is defined as $\lift{\textlog{lowloc}}{x}$, where
$\textlog{lowloc}$ is a predicate characterizing the region of memory containing
``low locations'' (distinct from the other region containing ``high
locations'').

@@

\newcommand{\OutV}{\textlog{Out}}

In Thomas' extension of OCPL, a new value $\OutV$ is added, denoting an
``output object capability'', as well as a $\textlog{print}$ primitive, where
$\textlog{print}\; \OutV\; v$ effectively ``prints'' the value $v$, i.e. adds it
to the trace of printed values.

@@

The theorem then becomes:
%
\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{\! P(t)}} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
    (C[e]);(\emptyset, \textlog{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \textlog{OK} \wedge P(t)}
\end{mathpar}

@@

To be slightly more general, we could actually allow sharing $\OutV$ in case the
predicate $P$ does not place any constraints on the output, i.e. it is the
$\TRUE$ predicate. Technically, we could hence have the following alternative
definition for $\textlog{lift}$ (although admittedly, it does look a bit
hard-coded):
%
\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
                            \pure{(\Lam{\_}.\TRUE) \spac t}}
  \end{array}
\]

@@

Similarly to $\textlog{lift}$, we can generalize the existing logical relation
to thread a constraint on directly accessible memory locations. One would
parameterize the value, expression and register relations ($\VR$, $\ER$ and $\RR$)
with a predicate $\Psi$ on addresses.

@@

Then, $\VR^{\notMMIO}$ is similar to the $\textlog{lowval}$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, nor can it gain access to memory-mapped locations indirectly,
similarly to how $\textlog{lowval}$s cannot grant access to high locations
directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
with values in $\VR^{\notMMIO}$, then the invariants in a private future world
of our starting world will be satisfied at the end of execution.

@@

Given this definition, we can now formalize the initialization constraints on
registers $r$ and memory $m$, given the region $\MMIO$ and
the driver's address space $[b,e)$ as follows (for a resource-aware formulation;
see the bootSpec postcondition below):

>>>>

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\X}[1]{\ensuremath{\mathrm{#1}}}
\newcommand{\V}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\I}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\Sf}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\T}[1]{\texttt{#1}}
\newcommand{\SL}{Separation Logic\xspace}
\newcommand{\pure}[1]{\tensor[^{\ulcorner}]{#1{}}{^{\urcorner}}} %Hacky {} to avoid double superscript

@@

\newcommand{\MMIO}{\Sf{MMIO}\xspace}

\DeclareMathOperator{\initOKo}{init_{OK}}
\DeclareMathOperator{\initOKos}{init^{\X{sem}}_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\newcommandx{\initOKS}[4][2=r,3=m,4=W]{\initOKos(#1,#2,#3,#4)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\DeclareMathOperator{\initBCos}{init^{\X{sem}}_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
\newcommandx{\initBCS}[3][1=r,2=m,3=W]{\initBCos(#1,#2,#3)}
\DeclareMathOperator{\driverC}{code}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\lowval}{\Sf{lowval}}
\DeclareMathOperator{\lifto}{\Sf{lift}}
\newcommandx{\trInv}[2][1=P,2=\iota]{\operatorname{trInv}^{#2}(#1)}
\newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}
\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
\newcommand{\app}{\mathbin{+\!\!+}}

@@

\begin{mathpar}
  \mprset{vskip=0.3em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \hoare{\TRUE}{e}{x\ldotp \lowval x} \\
    (C[e]);(\emptyset, \Sf{OK}) \longrightarrow^* T';(h',g')
  }
  {g' = \Sf{OK}}
\end{mathpar}

For any closed expression $e$, if $e$ has been verified to only return low
values, then running $e$ wrapped in an adversarial context $C$ from an initial
state, then we can observe that every reachable state is good
($g' = \Sf{OK}$ means that no assertion has failed in $e$).

$C \in AdvCtx$ means that $C$ cannot contain assertions (otherwise one could
trivially contradict the theorem by taking
$C[\cdot] = \Sf{assert}\spac \X{false}$), and cannot contain references to raw
memory locations (otherwise one could directly access $e$'s private state,
invalidating the local state encapsulation mechanisms).

\newcommand{\lift}[2]{\lifto #1 \: #2}
\newcommand{\liftP}[1]{\lift{\Psi}{#1}}

$\Sf{lowval}\; x$ intuitively means that $x$ cannot be used to directly
access private (or ``high'') locations. It is formally defined using a logical
relation ``$\liftP{v}$'', which more generally asserts that the value $v$ only
gives direct access to locations that satisfy the predicate $\Psi$.

\[
  \begin{array}{lcl}
    \liftP (\Sf{rec}\; f\; x.\; e) & \eqdef
    & \later \forall v.\; \hoare{\liftP{v}}{e[v/x,\Sf{rec}\;f\;x.\; e/f]}{y.\; \liftP y} \\
    %
    \liftP (v_{1}, v_{2}) & \eqdef & \later (\liftP v_{1}, \liftP v_{2}) \\
    %
    \color{BrickRed} \liftP \ell & \color{BrickRed} \eqdef & \color{BrickRed} \Psi \; \ell \\
    %
    \ldots
  \end{array}
\]

Then, $\lowval x$ is defined as $\lift{\Sf{lowloc}}{x}$, where
$\Sf{lowloc}$ is a predicate characterizing the region of memory containing
``low locations'' (distinct from the other region containing ``high
locations'').

@@

\newcommand{\OutV}{\Sf{Out}}

In Thomas' extension of OCPL, a new value $\OutV$ is added, denoting an
``output object capability'', as well as a $\Sf{print}$ primitive, where
$\Sf{print}\; \OutV\; v$ effectively ``prints'' the value $v$, i.e. adds it
to the trace of printed values.

@@

The theorem then becomes:
%
\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \trInv \vdash \hoare{\TRUE}{e}{x.\; \operatorname{\Sf{lowval}} x} \\
    (C[e]);(\emptyset, \Sf{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \Sf{OK} \wedge P(t)}
\end{mathpar}
, where we introduced the following notation:
%
\[
  \begin{array}{lcl}
    \trace(t) & \eqdef & \tracefrag{t}\\
    \trInv & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast \pure{\! P(t)}}
  \end{array}
\]

@@

To be slightly more general, we could actually allow sharing $\OutV$ in case the
predicate $P$ does not place any constraints on the output, i.e. it is the
$\TRUE$ predicate. Technically, we could hence have the following alternative
definition for $\Sf{lift}$ (although admittedly, it does look a bit
hard-coded):
%
\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \trInv[\Lam{\_}.\TRUE]
  \end{array}
\]

@@

Similarly to $\Sf{lift}$, we can generalize the existing logical relation
to thread a constraint on directly accessible memory locations. One would
parameterize the value, expression and register relations ($\VR$, $\ER$ and $\RR$)
with a predicate $\Psi$ on addresses.

@@

Then, $\VR^{\notMMIO}$ is similar to the $\lowval$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, nor can it gain access to memory-mapped locations indirectly,
similarly to how $\lowval$s cannot grant access to high locations
directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
with values in $\VR^{\notMMIO}$, then the invariants in a private future world
of our starting world will be satisfied at the end of execution.

@@

Given this definition, we can now formalize the initialization constraints on
registers $r$ and memory $m$, given the region \MMIO and
the driver's address space $[b,e)$ as follows (for a resource-aware formulation;
see the bootSpec postcondition below):
